Modal type theory based on the intuitionistic epistemic logic

### 2018-01-17 article.tex
added proof and erased some rubbish

  We build the smallest of subset of strongly normalazing terms of modal types and show that an arbitrary
  term belongs to this subset.

  \begin{defin} The set of strongly computable terms of type $\phi \in \mathbb{T}_{\textbf{K}}$, $SC_{\phi}$:

  \begin{itemize}
  \item Let $\phi = \textbf{K}\alpha$ and $\alpha \in \mathbb{T}$ , then:
  \begin{equation}
  SC_{\textbf{K}\alpha} = \{ M : \textbf{K}\alpha \: | \: \text{$M$ is strongly normalizing} \}
  \end{equation}

  \item Let $\phi = \textbf{K}(\tau \to \psi)$ and $\tau, \psi \in \mathbb{T}_{\textbf{K}}$, then:
  \begin{equation}
  SC_{\textbf{K}(\tau \to \psi)} = \{ M : \textbf{K}(\tau \to \psi) \: | \: \forall N \in SC_{\textbf{K}\tau}, M \star N \in SC_{\textbf{K}\psi}\}
  \end{equation}

  \item Let $\phi = \textbf{K}(\tau_1 \times \tau_2)$ and $\tau_1, \tau_2 \in \mathbb{T}_{\textbf{K}}$, then:
  \begin{equation}
  SC_{\textbf{K}(\tau_1 \times \tau_2)} = \{ P : \textbf{K}(\tau_1 \times \tau_2) \: |
  \: {\bf pure \:} (\lambda x. \pi_i x) \star P \in SC_{\textbf{K}\tau_i}, i \in \{1,2\} \}
  \end{equation}
  \end{itemize}
  \end{defin}

  If $M \in SC_{\alpha}$, then $M$ is strongly normalizing.

  1) If $M \in SC_{{\bf K} \alpha}$ and $\alpha \in \mathbb{T}$, then $M$ is strongly normalizing by the
  definition of $SC_{{\bf K} \alpha}$.

  2) Let $M \in SC_{\textbf{K}(\tau \to \psi)}$, so by every $N \in SC_{\textbf{K}\tau}$,
  $M \star N \in SC_{\textbf{K}\psi}$, which is strongly normalizing by hypothesis. So $M$ is
  strongly normalizing.

  3) Let $M \in SC_{\textbf{K}(\tau_1 \times \tau_2)}$, so ${\bf pure \:} (\lambda x. \pi_i x) \star M \in
  SC_{\textbf{K}\tau_i}$, $i \in \{ 1,2 \}$, which are strongly normalizing. So $M$ is
  strongly normalizing.

  Let $M \rightarrow_{\beta} M'$ and $M \in SC_{\alpha}$, then $M' \in SC_{\alpha}$.

  1) Let $M \rightarrow_{\beta} M'$ and $M \in SC_{\textbf{K}\alpha}$, where $\alpha \in \mathbb{T}$.

  $M$ has the longest reduction path (which we denote as $p(M)$). So $p(M') < p(M)$, then $M' \in SC_{\textbf{K}\alpha}$.

  2) Let $M \in SC_{\textbf{K}(\alpha \to \beta)}$ and $M \rightarrow_{\beta} M'$. Let $N \in
  SC_{\textbf{K}\alpha}$. So $M \star N \in SC_{\textbf{K}\beta}$.

  If $M \rightarrow_{\beta} M'$, then $M \star N \rightarrow_{\beta} M' \star N$ by reduction
  rule, so $M' \star N \in SC_{\textbf{K}\beta}$ and $M' \in SC_{\textbf{K}(\alpha \to \beta)}$ by
  hypothesis.

  3) Let $M \in SC_{\textbf{K}(\tau_1 \times \tau_2)}$ and $M \rightarrow_{\beta} M'$.

  So ${\bf pure \: } (\lambda x. \pi_i x) \star M \rightarrow_{\beta} {\bf pure \: } (\lambda x. \pi_i x)
  \star M'$, $i \in \{ 1, 2 \}$ by reduction rule. So ${\bf pure \: } (\lambda x. \pi_i x) \star M' \in
  SC_{\textbf{K} \tau_i }$ and $M' \in SC_{\textbf{K}(\tau_1 \times \tau_2)}$.

  \begin{defin} Neutral term:

  We define a term $M$ to be neutral if it has of the next forms:

  1) $M = x$, where $x \in \mathbb{V}$;

  2) $M = (P Q)$;

  3) $M = \pi_i M$, $i \in \{ 1,2 \}$;

  4) $M = P \star Q$;

  5) If $M$ is a neutral, then ${\bf pure \: } M$ is a neutral.

  Let $M \rightarrow_{\beta} M'$ and $M' \in SC_{\alpha}$ for every one-step reduction. So if $M'$ is a
  neutral, then $M \in SC_{\alpha}$.

  Simple induction on the structure of $M'$.

  Let $x_1 : \phi_1, \dots, x_n : \phi_n \vdash M : \phi$ and for all $i \in \{ 1, \dots, n\}$, $N_i
  \in SC_{\phi_i}$, then $(M [x_1 := N_1, \dots, x_n := N_n]) \in SC_{\phi}$.

  1) If $\phi$ is an atomic and $M$ is a variable, then this condition holds straightforwardly.

  2) Let $\Gamma = \{ x_1 : \phi_1, \dots, x_n : \phi_n \}$, $\Gamma \vdash {\bf pure \: } M :
  \textbf{K}\alpha$ and for all $i \in \{1, \dots, n\}, N_i \in SC_{\phi_i}$.

  Then by $\Gamma \vdash M : \alpha$ by generation
  and $(M [x_1 := N_1, \dots, x_n := N_n]) \in SC_{\alpha}$ by induction hypothesis.

  Hence, $\Gamma \vdash {\bf pure \: } M : \textbf{K}\alpha$ and $({\bf pure \: } M ([x_1 := N_1, \dots, x_n := N_n])) \in SC_{\textbf{K}\alpha}$ by definition of $SC_{\textbf{K}\alpha}$.

  3) Let $\Gamma = \{ x_1 : \phi_1, \dots, x_n : \phi_n \}$, $\Gamma : \phi_n \vdash M \star P : \textbf{K}\beta$ and forall $i \in \{ 1, \dots, n\}, N_i \in SC_{\phi_i}$.

  Then $\Gamma \vdash M : \textbf{K}(\alpha \to \beta)$, $\Gamma \vdash P : \textbf{K}\alpha$ by generation.

  But by induction hypothesis $M [x_1 := N_1, \dots, x_n := N_n] \in SC_{\textbf{K}(\alpha \to
  \beta)}$ and $P [x_1 := N_1, \dots, x_n := N_n] \in SC_{\textbf{K}\alpha}$.

  Then, by definition of $SC_{\textbf{K}\beta}$, $((M [x_1 := N_1, \dots, x_n := N_n]) \star (P [x_1 := N_1, \dots, x_n := N_n])) \in SC_{\textbf{K}\beta}$, i.e. $(M \star N ([x_1 := N_1, \dots, x_n := N_n])) \in
  SC_{\textbf{K}\beta}$.
  \end{proof}

  \begin{col}

  If $\vdash M : \alpha$, then $M$ is strongly normalizing.

  \end{col}

  $M \in SC_{\alpha}$ by Lemma 10, so $M$ is strongly normalizing.

@@

In the confluence proof (below) we treat the cases with ${\bf pure}$ and $\star$ similar to \cite{Baren} \cite{Baren2}.

\begin{defin} Alphabet for the labelled terms:

variables: $x, y, z, x_1, y_1, z_1, ...$;

lambdas: $\lambda, \lambda_0, \lambda_1, \lambda_2, ...$;

constructors for an applicative functor: ${\bf pure \: }, \star$;

parentheses (,).

\begin{defin} The set of labelled terms $\Lambda^{'}_{\textbf{K}}$ inductively defined as a set of words on the alphabet
described above:

1) $x \in \Lambda^{'}$;

2) If $M \in \Lambda^{'}_{\textbf{K}}$, then $(\lambda x. M) \in \Lambda^{'}_{\textbf{K}}$;

3) If $M, N \in \Lambda^{'}_{\textbf{K}}$, then $(MN) \in \Lambda^{'}_{\textbf{K}}$;

4) If $M \in \Lambda^{'}_{\textbf{K}}$, then ${\bf pure \: } M \in \Lambda^{'}_{\textbf{K}}$;

5) If $M, N \in \Lambda^{'}_{\textbf{K}}$, then $M \star N \in \Lambda^{'}_{\textbf{K}}$;

6) If $M, N \in \Lambda^{'}_{\textbf{K}}$, then for all $i \in \mathbb{N}$, $((\lambda_i x. M) N) \in
\Lambda^{'}_{\textbf{K}}$.

\begin{defin} Erasing map

Erasing map is a map $|.| : \Lambda^{'}_{\textbf{K}} \to \Lambda_{\textbf{K}}$, such that:

1) $|x| = x$;

2) $|(\lambda x. M)| = \lambda x. |M|$;

3) $|(M N)| = |M| |N|$;

4) $|({\bf pure \: } M)| = {\bf pure \: } |M|$;

5) $|M \star N| = |M| \star |N|$;

6) $|((\lambda_i x. M) N)| = (\lambda x. |M|) |N|$

\begin{ex}

$|{\bf pure \: } ((\lambda_i x. M) N) \star P| = {\bf pure \: } (\lambda x. |M|) |N|) \star |P|$

\end{ex}

\begin{defin} Substitution for $\Lambda^{'}_{\textbf{K}}$:

4) $({\bf pure \: } M) [x := P] = {\bf pure \: } \: (M [x := P])$;

5) $(M \star N) [x := P] = (M [x := P]) \star (N [x := P])$;

6) $(\lambda_i x. M) N [y := P] = (\lambda_i x. M [y := P]) (N [y := P])$.

\begin{defin} One-step reduction $\rightarrow_{\beta^{'}}$ for $\Lambda^{'}_{\textbf{K}}$:

1) $(\lambda x. M) N \rightarrow_{\beta^{'}} M [x := N]$;

2) ${\bf pure \: } (\lambda x. x) \star M \rightarrow_{\beta^{'}} M$;

3) ${\bf pure \: } (\lambda f g x. f (g x)) \star M \star N \star P \rightarrow_{\beta^{'}} M \star (N \star
P)$;

4) $({\bf pure \: } M) \star ({\bf pure \: } N) \rightarrow_{\beta^{'}} {\bf pure \: } (M N)$;

5) $M \star ({\bf pure \: } N) \rightarrow_{\beta^{'}} {\bf pure \: } (\lambda f. f N) \star M$;

6) $(\lambda_i x. M) N \rightarrow_{\beta^{'}} M [x := N]$.

Multi-step reduction $\twoheadrightarrow_{\beta^{'}}$ is a reflexive-transitive closure of
$\rightarrow_{\beta^{'}}$.

\begin{defin} Let us define a map $\phi : \Lambda^{'}_{\textbf{K}} \to \Lambda_{\textbf{K}}$ inductively as
follows:

1) $\phi(x) = x$;

2) $\phi(M N) = \phi(M)\phi(N)$;

3) $\phi(\lambda x. M) = \lambda x. \phi(M)$;

4) $\phi({\bf pure \: } M) = {\bf pure \: } (\phi(M))$;

5) $\phi(M \star N) = \phi(M) \star \phi(N)$;

6) $\phi((\lambda_i x. M) N) = \phi(M) [x := \phi(N)]$.

\begin{ex}

$\phi({\bf pure \: } ((\lambda_i x. M) N) \star P) = {\bf pure \: } (\phi(M) [x := \phi(N)]) \star \phi(P)$

\end{ex}

1) Let $M, N \in \Lambda^{'}_{\textbf{K}}$ and $|M| \twoheadrightarrow_{\beta} |N|$, then $M \twoheadrightarrow_{\beta^{'}} N$.

2) Let $M, N \in \Lambda^{'}_{\textbf{K}}$ and $M \twoheadrightarrow_{\beta^{'}} N$, then $|M| \twoheadrightarrow_{\beta} |N|$.

Induction on the generation of $\twoheadrightarrow_{\beta}$ ($\twoheadrightarrow_{\beta^{'}}$).

1) Let us consider homomorphism rule. The rest applicative reduction rules are considered similary.

Let $({\bf pure \: } M') \star ({\bf pure \: } N'), {\bf pure \: } (M' N') \in \Lambda^{'}_{\textbf{K}}$.

So $|({\bf pure \: } M') \star ({\bf pure \: } N')| = ({\bf pure \: } |M'|) \star ({\bf pure \: } |N'|)$ and
$|{\bf pure \: } (M' N')| = {\bf pure \: } (|M'| |N'|)$.

By reduction rule, $({\bf pure \: } |M'|) \star ({\bf pure \: } |N'|) \rightarrow_{\beta} {\bf pure \: }
(|M'||N'|)$.

But $({\bf pure \: } M') \star ({\bf pure \: } N') \rightarrow_{\beta^{'}} {\bf pure \: } (M' N')$ by
reduction rule for $\rightarrow_{\beta^{'}}$.

2) Let us consider interchange rule.

Let $M \star ({\bf pure \: } N), {\bf pure \: } (\lambda f. f N) \star M \in \Lambda^{'}_{\textbf{K}}$ and
$M \star ({\bf pure \: } N) \rightarrow_{\beta^{'}} {\bf pure \: } (\lambda f. f N) \star M$.

But $|M \star ({\bf pure \: } N)| = |M| \star ({\bf pure \: } |N|)$ and $|{\bf pure \: } (\lambda f. f N)
\star M| = {\bf pure \: } (\lambda f. f |N|) \star |M|$.

So $|M| \star ({\bf pure \: } |N|) \rightarrow_{\beta} {\bf pure \: } (\lambda f. f |N|) \star |M|$ by
$\beta$-reduction rule.

It is easy to see, that the statement for $\twoheadrightarrow_{\beta^{'}}$ and $\twoheadrightarrow_{\beta}$
immedeatly follows from transitivity of multi-step rediction for labelled terms and for usual terms
respectively.

$\phi(M [x := N]) = \phi(M) [x := \phi(N)]$.

\begin{proof} Induction on $M$.

1) Let $M = x$. Then $\phi(x [x := N]) = \phi(N)$.

On the other hand, $\phi(x) [x := \phi(N)] = x [x := \phi(N)] = \phi(N)$.

So $\phi(x [x := N]) = \phi(x) [x := \phi(N)]$.

2) Let $M = y$ and $y \neq x$. Then $\phi(y [x := N]) = \phi(y) = y$.

But $\phi(y) [x := \phi(N)] = y [x := \phi(N)] = y$.

Therefore $\phi(y [x := N]) = \phi(y) [x := \phi(N)]$.

3) Let $M = {\bf pure \: } M'$. Then $\phi({\bf pure \: } M' [x := N]) = {\bf pure \: } \phi(M' [x := N])$.

By hypothesis, ${\bf pure \: } (\phi(M' [x := N])) = {\bf pure \: } (\phi(M') [x := \phi(N)])$, which is
$({\bf pure \: } \phi(M')) [x := \phi(N)]$ by substitution definition.

4) Let $M = M' \star N'$. So $\phi((M' \star N') [x := N])) = \phi(M' [x := N] \star N' [x := N])$.

By definition of $\phi$,

\quad $\phi(M' [x := N] \star N' [x := N]) = \phi(M' [x := N]) \star \phi(N' [x := N])$.

But by induction hypothesis,

\quad $\phi(M' [x := N]) = \phi(M') [x := \phi(N)]$ and

\quad $\phi(N' [x := N]) = \phi(N') [x := \phi(N)]$.

Hence,

\quad $\phi(M' [x := N]) \star \phi(N' [x := N]) = \phi(M') [x := \phi(N)] \star \phi(N') [x := \phi(N)]$.

So,

\quad $\phi(M') [x := \phi(N)] \star \phi(N')[x := \phi(N)] = (\phi(M') \star \phi(N')) [x := \phi(N)]$.

And by definition of $\phi$, $(\phi(M') \star \phi(N')) [x := \phi(N)] = \phi(M' \star N') [x := \phi(N)]$.

  Let $M, N \in \Lambda^{'}_{\textbf{K}}$ and $M \twoheadrightarrow_{\beta^{'}} N$, then $\phi(M)
  \twoheadrightarrow_{\beta} \phi(N)$.

1) Let ${\bf pure \: } (\lambda x. x) \star M, M \in \Lambda^{'}_{\textbf{K}}$ and ${\bf pure \: } (\lambda
x. x) \star M \rightarrow_{\beta^{'}} M$.

But $\phi ({\bf pure \: } (\lambda x. x) \star M) = {\bf pure \: } (\lambda x. x) \star \phi(M)$.\

So ${\bf pure \: } (\lambda x. x) \star \phi(M) \rightarrow_{\beta} \phi(M)$ by $\beta$-reduction rule.

2) Let ${\bf pure \: } (\lambda f g x. f (g x)) \star M \star N \star P, M \star (N \star P) \in
\Lambda^{'}_{\textbf{K}}$ and ${\bf pure \: } (\lambda f g x. f (g x)) \star M \star N \star P
\rightarrow_{\beta^{'}} M \star (N \star P)$.

By the definition of $\phi$:

$\phi({\bf pure \: } (\lambda f g x. f (g x)) \star M \star N \star P) = {\bf pure \: } (\lambda f g x. f (g
x)) \star \phi(M) \star \phi(N) \star \phi(P)$;

$M \star (N \star P) = \phi(M) \star (\phi(N) \star \phi(P))$.

Hence, ${\bf pure \: } (\lambda f g x. f (g x))
\star \phi(M) \star \phi(N) \star \phi(P) \rightarrow_{\beta} \phi(M) \star (\phi(N) \star \phi(P))$ by
$\beta$-reduction rule.

3) Let $({\bf pure \: } M) \star ({\bf pure \: } N), {\bf pure \: } (M N) \in \Lambda^{'}_{\textbf{K}}$ and
$({\bf pure \: } M) \star ({\bf pure \: } N) \rightarrow_{\beta} {\bf pure \: } (M N)$.

By the definition of $\phi$:

$\phi(({\bf pure \: } M) \star ({\bf pure \: } N)) = ({\bf pure \: } \phi(M)) \star ({\bf pure \: }
\phi(N))$;

$\phi({\bf pure \: } (M N)) = {\bf pure \: } (\phi(M) \phi(N))$.

So, by reduction rule, $({\bf pure \: } \phi(M)) \star ({\bf pure \: } \phi(N)) \rightarrow_{\beta} {\bf
pure \: } (\phi(M) \phi(N))$.

4) Let $M \star ({\bf pure \: }), {\bf pure \: } (\lambda f. f N) \star M$ and $M \star ({\bf pure \: } N)
\rightarrow_{\beta^{'}} (\lambda f. f N) \star M$.

$\phi(M \star ({\bf pure \: } N)) = \phi(M) \star ({\bf pure \: } \phi(N))$

$\phi((\lambda f. f N) \star M) = (\lambda f. f \phi(N)) \star \phi(M)$.

So, $\phi(M) \star ({\bf pure \: } \phi(N)) \rightarrow_{\beta} {\bf pure \: } (\lambda f. f \phi(N)) \star
\phi(M)$.
\end{proof}

Let $M \in \Lambda^{'}_{\textbf{K}}$. Then $|M| \twoheadrightarrow_{\beta} \phi(M)$.

Induction on the structure of $M$.

\begin{lemma} Strip lemma.

If $M \rightarrow_{\beta} N$ and $M \twoheadrightarrow_{\beta} P$. Then there exists some term $Q$, such that
$N \twoheadrightarrow_{\beta} Q$ and $P \twoheadrightarrow_{\beta} Q$.

  Proof is similar to \cite{Baren} \cite{Baren2}. We build the following diagram

\xymatrix{
&&& M \ar[dd]_{\beta} \ar@{->>}[rrr] &&& P \ar@{-->>}[dd]^{\beta} \\
&&&& N' \ar[ul]^{|.|} \ar[dl]_{\phi} \ar@{-->>}[rrr]^{\beta^{'}} &&& P' \ar[ul]^{|.|} \ar[dl]_{\phi} \\
&&& N \ar@{-->>}[rrr]_{\beta} &&& Q
}

which is commutes by lemmas 11 -- 14.

\begin{theorem} Confluence.

If $M \twoheadrightarrow_{\beta} N$ and $M \twoheadrightarrow_{\beta} P$. Then there exists some term $Q$,
such that $N \twoheadrightarrow_{\beta} Q$ and $P \twoheadrightarrow_{\beta} Q$.

  By unfolding $M \twoheadrightarrow_{\beta} N$ as the sequence of one-step reductions $M \rightarrow_{\beta}
  M_1 \rightarrow_{\beta} ... \rightarrow_{\beta} M_n \rightarrow_{\beta} N$ and applying strip lemma on every step.

@@

@@

@@

  \bibitem{ModalK} Yoshihiko KAKUTANI, A Curry-Howard Correspondence for Intuitionistic Normal Modal Logic, Computer Software, Released February 29, 2008, Online ISSN , Print ISSN 0289-6540.
  \end{thebibliography}

>>>>

@@

@@

  i) Let $\Gamma \vdash M : A$ and $M \twoheadrightarrow_{\beta} N$, then $[\![\Gamma \vdash M : A]\!] = [\![\Gamma \vdash N : A]\!]$;

  ii) Let $\Gamma \vdash M : A$ and $M \twoheadrightarrow_{\eta} N$, then $[\![\Gamma \vdash M : A]\!] = [\![\Gamma \vdash N : A]\!]$;
\end{lemma}

\begin{proof}
  i) For $\beta$-reduction

@@

Cases with $\beta$-reductions for $let_{\bf K}$ are shown in \cite{ModalK1}. Let us consider cases with ${\bf pure}$.

1) $[\![\Gamma \vdash {\bf pure \:} ((\lambda x. M) N) : {\bf K}B]\!] = [\![\Gamma \vdash {\bf pure \:} (M [x := N]]) : {\bf K}B]\!]$

$\begin{array}{lll}
&[\![\Gamma \vdash {\bf pure \:} (\lambda x. M) N : {\bf K}B]\!] = & \text{By interpretation} \\
&\eta_{[\![B]\!]} \circ (\epsilon \circ \langle \Lambda([\![M]\!]), [\![N]\!]\rangle) =& \text{Property of $\times$} \\
&\eta_{[\![B]\!]} \circ (\epsilon \circ (\Lambda([\![M]\!]) \times id_{[\![A]\!]}) \circ \langle id_{[\![\Gamma]\!]}, [\![N]\!] \rangle) = & \text{Associativity of composition} \\
&\eta_{[\![B]\!]} \circ ((\epsilon \circ (\Lambda([\![M]\!]) \times id_{[\![A]\!]})) \circ \langle id_{[\![\Gamma]\!]}, [\![N]\!] \rangle) = & \text{Exponentiation property} \\
&\eta_{[\![B]\!]} \circ ([\![M]\!] \circ \langle id_{[\![\Gamma]\!]}, [\![N]\!] \rangle) = & \text{Substitution lemma} \\
&\eta_{[\![B]\!]} \circ [\![M [\vec{x}, x := \vec{x}, N]]\!] =& \text{By interpretation} \\
&[\![\Gamma \vdash {\bf pure \:} (M [x := N]]) : {\bf K}B]\!]&
\end{array}$

2) $[\![\Gamma \vdash {\bf pure \:} (\pi_i \langle [\![M_1]\!], [\![M_2]\!] \rangle) : {\bf K}A_i ]\!] = [\![\Gamma \vdash {\bf \: pure} M_i : {\bf K}A_i ]\!]$

$\begin{array}{lll}
&[\![\Gamma \vdash {\bf pure \:} (\pi_i \langle M_1, M_2] \rangle) : {\bf K}A_i ]\!] = & \text{By interpretation} \\
&\eta_{[\![A_i]\!]} \circ \pi_i \circ \langle [\![M_1]\!], [\![M_2]\!] \rangle = & \text{Property of $\times$} \\
&\eta_{[\![A_i]\!]} \circ [\![M_i]\!] =& \text{By interpetation} \\
&[\![\Gamma \vdash {\bf pure \:} M_i : {\bf K}A_i]\!]&
\end{array}$

@@

  \bibitem{ModalK} Yoshihiko KAKUTANI, A Curry-Howard Correspondence for Intuitionistic Normal Modal Logic, Computer Software, Released February 29, 2008, Online ISSN , Print ISSN 0289-6540.

    \bibitem{ModalK1} Kakutani Y. (2007) Call-by-Name and Call-by-Value in Normal Modal Logic. In: Shao Z. (eds) Programming Languages and Systems. APLAS 2007. Lecture Notes in Computer Science, vol 4807. Springer, Berlin, Heidelberg
  \end{thebibliography}
